```{r options, echo = FALSE}
library(knitcitations)
```

# Introducing the R **gimms** package

### What it is all about
We've been collecting functions related to the download and processing of AVHRR 
GIMMS data for quite some time and with the most recent update to GIMMS3g 
`r citep("10.3390/rs6086929")`, we thought it was a good time to stuff the most 
fundamental work steps into a proper R package. In this context, 'most 
fundamental' refers to certain operations which we tended to repeat over and 
over again, including

* list all GIMMS files available online at [NASA ECOCAST](http://ecocast.arc.nasa.gov/data/pub/gimms/3g.v0/),
* download selected (or all) files, 
* re-arrange the list of downloaded files according to date, 
* transform binary data in ENVI format to proper **raster** format `r citep(citation("raster"))` and 
* aggregate the bi-monthly datasets to monthly maximum value composites (MVC).

In the following, you'll find a short introduction of what we came up with so 
far. Feel free to comment, raise issues and provide (constructive) criticism.
Any suggestions on how to improve the **gimms** package are highly appreciated!

### How to install
So far, the **gimms** package has not been submitted to CRAN but a preliminary 
package version can be installed directly from 
[GitHub](https://github.com/environmentalinformatics-marburg/gimms) via 

```{r install, message = FALSE}
# ## install 'gimms' package
# library(devtools)
# install_github("environmentalinformatics-marburg/gimms")

## load 'gimms' package
library(gimms)
```

There is no such thing as a 'development' branch yet, but be assured that such 
things will possibly be provided in the near-distant future given that there is 
a need for further functionality, improvements, bug-fixes etc. 

### List available files
For any further processing, it is helpful to know which GIMMS files are 
currently hosted on the [ECOCAST servers]((http://ecocast.arc.nasa.gov/data/pub/gimms/3g.v0/). 
`updateInventory` has been designed just for that purpose as it imports the file 
inventory stored online in *00FILE-LIST.txt* as 'character' vector directly into 
R. By setting `sort = TRUE`, it is even possible to return the output sorted by 
date which is anything but intuitive when dealing with naming conventions in the 
form of 'geo83sep15a.n07-VI3g'). If there is no active internet connection 
available, `updateInventory` automatically imports the latest offline version of 
the file inventory which is stored (and regularly updated) in 
'inst/extdata/inventory.rds'.

```{r updateInventory}
gimms_files <- updateInventory(sort = TRUE)
gimms_files[1:10]
```

### Download files
With the information on files currently hosted on the remote server at hand, the 
next logical step of the **gimms** processing chain is to download selected (if 
not all) bi-monthly datasets. This can be achieved by running `downloadGimms`, 
but since the function works with different types of input parameter (if any), 
some *ex ante* information are possibly helpful for its proper use.
further explanation seems reasonable.
